### 2021-03-19 newvariant.tex
control seq typos!
some more random comments

\newcommand{\rx}[1]{\ensuremath{{r}_{#1}}\xspace} 
\newcommand{\ly}[1]{\lx{\mathrm{#1}}}

@@

We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and a focal variant of concern (e.g., B.1.1.7).
Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by the wild type $w$ and the variant $v$) can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
\begin{equation}
i_x(t) = \int_0^\infty i_x(t-\tau) K_x(t, \tau) \dtau.
\end{equation}
This framework provides a flexible way of modeling disease dynamics and generalizes compartmental models, such as the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, roberts2007model, champredon2018equivalence}.

The integral of the kernel $\RR_x(t) = \int K_x(t, \tau) \dtau$---referred to as the instantaneous reproduction number---describes the average infectiousness of previously infected individuals at time $t$ \citep{fraser2007estimating}.
The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---which we refer to as the instantaneous generation-interval distribution---describes their relative contribution to current incidence $i_x(t)$ and provides information about the time scale of disease transmission.
Both the reproduction number and the generation-interval distribution can depend on several factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility \citep{fraser2007estimating}.

Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
In this case, the incidence of either strains can exhibit exponential growth or decay at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
\begin{equation}
\frac{1}{\RR_x} = \int_0^\infty \exp(- r_x \tau) g_x(\tau) \dtau.
\end{equation}
We can approximate the $r$--$\RR$ relationship by assuming that the generation-interval distribution is approximately gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$ \citep{park2019practical}:
\begin{equation}
\RR_x \approx (1 + \kappa_x r_x \bar{G}_x)^{1/\kappa_x}.
\end{equation}
The gamma assumption covers models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}; various gamma assumptions and approximations are widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.
For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_w = \kappa_v = \kappa$; instead, we focus on the effect of potential differences in the mean generation intervals.

@@

Epidemic speed $r_x$ is often estimated directly from case data \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}.
Studies of new SARS-CoV-2 variants, however, have mostly focused on characterizing changes in \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
This has the advantage of correcting for testing patterns and intensity, and possibly for other transient effects that might affect variants and wild type viruses similarly.
When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic equation:
\begin{align}
p(t) &= \frac{i_v(t_0) \exp(r_v t)}{i_w(t_0) \exp(r_w t) + i_v(t_0) \exp(r_v t)},
\\ &= \frac{1}{1 + \left(i_w(t_0)/i_v(t_0)\right) \exp(-\delta t)},
\end{align}
where the logistic growth rate $\delta = r_v - r_w$ corresponds to the relative speed of the epidemic.

@@

As our baseline scenario, we assume $\delta = 0.1/\textrm{days}$ based on the observed relative speed of the new variant in the UK \citep{davies2021estimated}.
This speed can be then used to estimate the relative strength $\theta = \RR_v/\RR_w$ of the new variant with a caveat; such inferece depends on assumptions about the generation-interval distribution.
The mean generation interval $\bar{G}_w$ of the wild type is assumed to equal 5 days \citep{ferretti2020quantifying}.
The squared coefficient of variation $\kappa$ in generation intervals os assumed to equal 0.2 \citep{ferretti2020quantifying}---for simplicity, we assume that both the variant and the wild type have equal $\kappa$ and only consider differences in the mean.
As we expect there to be greater uncertainty in $\kappa$ than in $\bar{G}$, we evaluate the estimates of relative strength $\theta$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution).

>>>>

\newcommand{\rx}[1]{\ensuremath{{r}_{#1}}\xspace} 
\newcommand{\ry}[1]{\rx{\mathrm{#1}}}

@@

We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and a focal variant of concern (e.g., B.1.1.7).
Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by the wild type $w$ and the variant $v$ can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
\begin{equation}
i_x(t) = \int_0^\infty i_x(t-\tau) K_x(t, \tau) \dtau.
\end{equation}
This framework provides a flexible way of modeling disease dynamics and generalizes compartmental models, such as the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, roberts2007model, champredon2018equivalence}.

The integral of the kernel $\RR_x(t) = \int K_x(t, \tau) \dtau$---referred to as the instantaneous reproduction number---describes the average infectiousness of previously infected individuals at time $t$ \citep{fraser2007estimating}.
\jd{This is a particular kind of weighted average – in particular, it's not weighted by the actual number of infected individuals present. I kind of want to say GI-weighted average, but then we would have to introduce GI as an abbrv.}
The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---which we refer to as the instantaneous generation-interval distribution---describes their relative contribution to current incidence $i_x(t)$ and provides information about the time scale of disease transmission.
Both the reproduction number and the generation-interval distribution can depend on several factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility \citep{fraser2007estimating}.

Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
In this case, the incidence of each strain will exhibit exponential growth (or decay) at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
\begin{equation}
\frac{1}{\RR_x} = \int_0^\infty \exp(- r_x \tau) g_x(\tau) \dtau.
\end{equation}
We can approximate the $r$--$\RR$ relationship by assuming that the generation-interval distribution is approximately gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$ \citep{park2019practical}:
\begin{equation}
\RR_x \approx (1 + \kappa_x r_x \bar{G}_x)^{1/\kappa_x}.
\end{equation}
The gamma assumption covers models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}; various gamma assumptions and approximations are widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.
For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_w = \kappa_v = \kappa$; instead, we focus on the effect of potential differences in the mean generation intervals.

@@

Epidemic speed $r_x$ is often estimated directly from case data \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}.
Studies of new SARS-CoV-2 variants, however, have mostly focused on characterizing changes in \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
\jd{I probably added patterns below, but now I want to drop it. Sometimes the testing pattern is: we just noticed the new variant and so let's go and test intensively in the communities where we think it is.}
This has the advantage of correcting for testing patterns and intensity, and possibly for other transient effects that might affect variants and wild type viruses similarly.
When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic equation:
\begin{align}
p(t) &= \frac{i_v(t_0) \exp(r_v t)}{i_w(t_0) \exp(r_w t) + i_v(t_0) \exp(r_v t)},
\\ &= \frac{1}{1 + \left(i_w(t_0)/i_v(t_0)\right) \exp(-\delta t)},
\end{align}
where the logistic growth rate $\delta = r_v - r_w$ corresponds to the relative speed of the epidemic.

@@

As our baseline scenario, we assume $\delta = 0.1/\textrm{days}$ based on the observed relative speed of the new variant in the UK \citep{davies2021estimated}.
This speed can be then used to estimate the relative strength $\theta = \RR_v/\RR_w$ of the new variant with a caveat; such inference depends on assumptions about the generation-interval distribution.
The mean generation interval $\bar{G}_w$ of the wild type is assumed to equal 5 days \citep{ferretti2020quantifying}.
The squared coefficient of variation $\kappa$ in generation intervals os assumed to equal 0.2 \citep{ferretti2020quantifying}---for simplicity, we assume that both the variant and the wild type have equal $\kappa$ and only consider differences in the mean.
As we expect there to be greater uncertainty in $\kappa$ than in $\bar{G}$, we evaluate the estimates of relative strength $\theta$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution).
